---
title: "Faithful visualisation with vanilla R"
author: "Dylan Chua"
date: "October 2024"
output:
  html_document:
    self_contained: yes
---

```{r setup-chunk, include=FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      cache = TRUE,
                      fig.path = "plots/R/")
```


# 1 Importing data and summary statistics
```{r import}
library(readr)
faithful <- read_csv("faithful.csv",
  col_types = cols(
    waiting = col_integer(),
    duration = col_double(),
    day = col_integer()
  )
)

# waiting time between successive eruptions
waiting <- faithful$waiting
# duration of the eruption
duration <- faithful$duration
# day recorded
day <- faithful$day
```

```{r summary}
# summary statistics of waiting time
summary(waiting)
# summary statistics of duration
summary(duration)
```

## boxplot of waiting time against day
```{r boxplot}
library(ggplot2)
ggplot(faithful) +
  aes(x = day, y = waiting, group = day) +
  geom_boxplot(fill = "orange") +
  labs(x = "Day", y = "Waiting time between successive eruptions / min") +
  theme_bw()
```

# 2 Histograms and kernel density plots

## histogram of waiting time with density plot
```{r histogram}
# number of bins for histogram
bins <- c(10, 25, 50)

library(ggpubr)

histograms <- lapply(
  bins,
  function(b) {
    ggplot(faithful) +
      aes(x = waiting) +
      geom_histogram(
        aes(y = after_stat(density)),
        binwidth = (max(waiting) - min(waiting) + 1) / b,
        fill = "#0088ff",
        colour = "black"
      ) +
      geom_density(
        lwd = 2,
        colour = "red"
      ) +
      labs(x = "Waiting time / min") +
      theme_bw()
  }
)
ggarrange(plotlist = histograms, ncol = 3)
```

## density plot of waiting time with different bandwidths
```{r densities}
colours <- c("bw = default" = "red", "bw = 2" = "green", "bw = 0.75" = "blue")

ggplot(faithful) +
  aes(x = waiting, color = "bw = default") +
  geom_density(
    lwd = 2,
  ) +
  geom_density(
    aes(x = waiting, color = "bw = 2"),
    bw = 2,
    lwd = 2,
    linetype = "dashed",
  ) +
  geom_density(
    aes(x = waiting, color = "bw = 0.75"),
    bw = 0.75,
    lwd = 2,
    linetype = "dotted",
  ) +
  labs(x = "Waiting time / min", color = "Legend") +
  theme_bw() +
  scale_color_manual(values = colours)
```

# 3 Plotting consecutive eruption waiting times

```{r scatter}
# x-axis limits: 0 to max number of eruptions in a day
xlim <- c(0, max(table(day)))
# y-axis limits: min to max waiting time
ylim <- c(min(waiting), max(waiting))

scatters <- lapply(
  1:15,
  function(i) {
    select <- which(day == i)
    ggplot(faithful[select, ]) +
      aes(x = seq_along(waiting), y = waiting) +
      geom_line(colour = "blue") +
      labs(
        title = paste("Day", i),
        x = "Eruption number",
        y = "Waiting time / min"
      ) +
      theme_bw() +
      coord_cartesian(xlim = xlim, ylim = ylim)
  }
)

ggarrange(plotlist = scatters, ncol = 5, nrow = 3)
```

# 4 Scatterplots and linear regression

# scatterplot of waiting time against lagged duration
```{r scatterplot}
n <- length(waiting)
library(tsutils)
lagduration <- lagmatrix(duration, 1)

ggplot(faithful) +
  aes(x = lagduration, y = waiting) +
  geom_point(
    colour = "orange",
  ) +
  labs(x = "Duration of previous eruption / min", y = "Waiting time / min") +
  theme_bw() +
  # linear regression of waiting time against lagged duration
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE)
```

# 5 K-means clustering

```{r k-2}
k <- 2
c <- kmeans(cbind(lagduration[2:n], waiting[2:n]), k)
lagdata <- data.frame(lagduration = lagduration[2:n], waiting = waiting[2:n])
ggplot(lagdata) +
  aes(x = lagduration, y = waiting, colour = factor(c$cluster)) +
  geom_point() +
  labs(
    x = "Duration of previous eruption / min",
    y = "Waiting time / min",
    colour = "Cluster"
  ) +
  theme_bw()
```

# different number of clusters
```{r k-3}
k <- 3
c <- kmeans(cbind(lagduration[2:n], waiting[2:n]), k)
ggplot(lagdata) +
  aes(x = lagduration, y = waiting, colour = factor(c$cluster)) +
  geom_point() +
  labs(
    x = "Duration of previous eruption / min",
    y = "Waiting time / min",
    colour = "Cluster"
  ) +
  theme_bw()
```

# lagged waiting time against waiting time
```{r k-lag}
lagwaiting <- lagmatrix(waiting, 1)
k <- 2
c <- kmeans(cbind(lagwaiting[2:n], waiting[2:n]), k)
lagdata <- data.frame(lagwaiting = lagwaiting[2:n], waiting = waiting[2:n])
ggplot(lagdata) +
  aes(x = lagwaiting, y = waiting, colour = factor(c$cluster)) +
  geom_point() +
  labs(
    x = "Waiting time of previous eruption / min",
    y = "Waiting time / min",
    colour = "Cluster"
  ) +
  theme_bw()
```
