---
title: "MECH60017 STATISTICS EXERCISE SHEET 1"
author: "Dylan Chua"
date: "October 2024"
output:
  html_document:
    self_contained: yes
---

```{r setup-chunk, include=FALSE}
knitr::opts_chunk$set(
  dev = "png",
  dpi = 300,
  cache = TRUE,
  fig.path = "plots/R/"
)
```

A Geiger counter was used to record the number of alpha particles emitted by a radioactive material in 1 minute intervals and the results are displayed below.
You will need to use this dataset for questions 1 and 2.

|    |    |    |    |    |    |    |    |    |    |
|:--:|:--:|:--:|:--:|:--:|:--:|:--:|:--:|:--:|:--:|
| 15 | 25 | 22 | 31 | 25 | 19 | 8  | 24 | 44 | 30 |
| 34 | 12 | 7  | 33 | 19 | 20 | 19 | 42 | 38 | 27 |
|    |    |    |    |    |    |    |    |    |    |

```{r geiger}
geiger <- c(
  15, 25, 22, 31, 25, 19, 8, 24, 44, 30,
  34, 12, 7, 33, 19, 20, 19, 42, 38, 27
)
```

## Q1
Construct a suitable histogram plot for the Geiger counter data.
---

```{r histogram}
library(ggplot2)

ggplot() +
  aes(x = geiger) +
  geom_histogram(binwidth = 8, fill = "orange", color = "black") +
  labs(x = "Number of alpha particles emitted in 1 minute", y = "Frequency") +
  theme_bw()
```

## Q2
Determine the sample mean, the sample variance, the median and the mode of the Geiger
counter data.
---

```{r summary}
# sample mean
mean(geiger)
# sample variance
var(geiger)
# median
median(geiger)
# mode
modes <- function(x) {
  ux <- unique(x)
  tab <- tabulate(match(x, ux))
  ux[tab == max(tab)]
}
modes(geiger)
```

## Q3
A random sample of 60 observations is such that $\sum_i x_i = 377$ and $\sum_i x_i^2 = 2377$.
Starting from the definition of the sample variance:
$$
s^2 = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2
$$
where $\bar{x}$ is the sample mean.

Show that the variance of the sample can be equivalently written as:
$$
s^2 = \frac{1}{n-1} \left[ \left(\sum_{i=1}^n x_i^2 \right) - n \bar{x}^2 \right]
$$

Hence, find the sample mean and standard deviation of this dataset.
---

**Solution:**

$$
s^2 = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2 \\
    = \frac{1}{n-1} \sum_{i=1}^n (x_i^2 - 2x_i\bar{x} + \bar{x}^2) \\
    = \frac{1}{n-1} \left[ \left(\sum_{i=1}^n x_i^2  \right) - 2\bar{x} \left(\sum_{i=1}^n x_i \right) + n\bar{x}^2 \right] \\
    = \frac{1}{n-1} \left[ \left(\sum_{i=1}^n x_i^2  \right) - 2n\bar{x}^2 + n\bar{x}^2 \right] \\
    = \frac{1}{n-1} \left[ \left(\sum_{i=1}^n x_i^2  \right) - n\bar{x}^2 \right]
$$

```{r sample}
# sample size
n <- 60
# sum of observations
sum_x <- 377
# sum of squares of observations
sum_x2 <- 2377

# sample mean
sample_mean <- sum_x / n
sample_mean
# standard deviation
sample_sd <- sqrt((sum_x2 - n * sample_mean^2) / (n - 1))
sample_sd
```

## Q4
A frequency table for student performances is as follows:

|Class  | Dash list             | Frequency | Proportion (%)|
|:-----:|:----------------------|:---------:|:-------------:|
|19-20  | //                    | 2         | 3.51          |
|20-21  | /                     | 1         | 1.75          |
|21-22  | ///// ////            | 9         | 15.79         |
|22-23  | ///// ///// ///// //  | 17        | 29.82         |
|23-24  | ///// ///// /         | 11        | 19.30         |
|24-25  | ///// //              | 7         | 12.28         |
|25-26  | /////                 | 5         | 8.77          |
|26-27  | /                     | 1         | 1.75          |
|27-28  | ///                   | 3         | 5.26          |
|28-29  | /                     | 1         | 1.75          |
|       |                       |           |               |
|**Sum**|                       | 57        | 99.98         |
|       |                       |           |               |

```{r performance}
# class intervals
class_intervals <- c(19, 20, 21, 22, 23, 24, 25, 26, 27, 28)
# frequencies
frequencies <- c(2, 1, 9, 17, 11, 7, 5, 1, 3, 1)
# number of observations
n <- sum(frequencies)
```

(a) Suppose we want to estimate the sample mean of this dataset by assuming that any point in the 19-20 range is taken to have value 19.5, any point in the 20-21 range has value 20.5, etc. Using this approach estimate the sample mean and variance of the strength dataset.

```{r strength}
# midpoints of class intervals
classes <- class_intervals + 0.5
classes
# sample mean
mean <- sum(classes * frequencies) / n
mean
# sample variance
variance <- sum((classes - mean)^2 * frequencies) / (n - 1)
variance
```

(b) Estimate the minimum and maximum values of the sample mean.

```{r min-max}
# minimum value of sample mean
classes <- class_intervals
min_mean <- sum(classes * frequencies) / n
min_mean
# maximum value of sample mean
classes <- class_intervals + 1
max_mean <- sum(classes * frequencies) / n
max_mean
```

which makes sense since we are shifting the class values by $\pm 0.5$ for all observations, hence the minimum and maximum values of the sample mean will be shifted by $\pm 0.5$ as well.